Motional Quantum Ground-State Cooing of a Single Sodium Atom

### 2017-06-19 paper.tex
Number update; Remove almost all "?"

\begin{figure*}
  \includegraphics[height=5.2cm]{imgs/Na_RSC_schematic.pdf}
  \includegraphics[height=4.5cm]{sequence.pdf}
  \caption{(A) Energy levels and schematics of Raman sideband cooling.
    The Raman transitions has a one photon detuning $\Delta=25GHz$ from the D2 line.
    We use D1 light with $\sigma^-$ polarization to repump atom out of $|F=2,m_F=-1\rangle$
    state to minimize heating on the atom in $|F=2,m_F=-2\rangle$ state.
    (B) Geometry and polarizations of the Raman and optical pumping beams relative to the
    optical tweezer and bias magnetic field.
    (C) Schematics of the cooling sequence. The tweezer is switching at $3\text{MHz}$ to
    reduce light shift during optical pumping. Each cooling cycle consists of $8$ pulses.
    The four axial pulses are addressing two neighboring cooling orders.
    The two pulses in each radial directions are either addressing two neighboring cooling orders
    or having different length on the first order when most of the population are below $n=3$
    towards then end of the cooling sequence.
    \label{f-setup}}
\end{figure*}
\begin{figure}[b]
  \includegraphics[width=8.5cm]{imgs/fig2_raman_op.pdf}
  \caption{Matrix elements and heating probabilities as a function of motional state.
    The range plotted covers $99\%$ of the initial thermal distribution.
    (A) Matrix elements for Raman transition in the axial direction showing deviation from
    $\sqrt{n}$ scaling and multiple minimums for different sideband orders.
    (B) Heating probability during the optical pumping step for all three axis.
    Due to the large Lamb-Dicke parameter,
    there is a high probability of heating especially in the axial direction.
    \label{f-ld}}
\end{figure}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_r.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_p1.pdf}
  \caption{Radial Raman sideband spectrum of first order heating, first order cooling and
    second order cooling (A) before and (B) after Raman sideband cooling.
    Rabi flopping on axis 3 (C) carrier and (D) first order heating sideband
    after cooling.
    Solid lines in (C) and (D) are theoretical fit with a ground state probability of $????\%$.
    Horizontal dashed lines in (A) to (D) are total survival and detection efficiency.
    \label{f-radial}}
\end{figure*}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_a1.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_p1.pdf}
  \caption{Axial Raman sideband spectrum from first order heating to eighth order cooling
    (A) before and (B) after Raman sideband cooling.
    The data for the second and higher orders of cooling sidebands are taken with $150\mu s$
    pulse time and the rest are taken with $125\mu s$ pulse time.
    Rabi flopping on axial (C) carrier and (D) first order heating sideband
    after cooling.
    Solid lines in (C) and (D) are theoretical fit with a ground state probability of $????\%$.
    Horizontal dashed lines in (C) and (D) are total survival and detection efficiency.
    \label{f-axial}}
\end{figure*}

@@

The Raman sideband cooling we used to achieve the high ground state preparation fidelity
consists of multiple cycles of laser pulses to manipulate the internal and
motional states of the atoms.
Figure \ref{f-setup}A shows the schematic of the energy levels and the cooling sequence
in our setup.
Each cooling cycle starts with the sodium atom in the $|F=2, m_F=-2\rangle$
ground electronic state and a certain vibration state $n$.
In the first step, a Raman pulse drives a transition to the motional state $n-\Delta n$
to reduce the motional energy while also changing the internal state to $|F=1, m_F=-1\rangle$.
In the second step, which finishes the cooling cycle,
an optical pumping pulse bring the atom back to the $|F=2, m_F=-2\rangle$ state to take away
the entropy. The second step could also change the motional state of the atoms which
could cause heating. The possibility for this to happen for an atom in the motional level $n$
is approximately proportional to the effective Lamb-Dicke parameter
$\eta^{OP}_{eff}=\sqrt{2n+1}\eta^{OP}$ where $\eta^{OP}=k^{OP}???$ is the Lamb-Dicke parameter for
optical pumping.
The geometry of the relevant beams and their polarizations is showed in figure \ref{f-setup}B.
The optical tweezer has one weakly confined axial direction (axis $1$) and
two more strongly confined radial directions (axis $2$ and $3$).
Multiple pairs of beams are used to drive the Raman transition during cooling in order to
isolate and maximize the coupling to different trap axis.

Although Raman sideband cooling to motional ground state has already been successfully
implemented on neutral atoms in other experiments using heavier species like Rubidium
\cite{Thompson2013,Kaufman2012}
and Cesium [????], such experiments generally use good polarization gradient cooling
and a tight confinement to achieve a small Lamb-Dicke parameter and effective Lamb-Dicke parameter.
However, this regime is harder to achieve with sodium atoms, creating additional challenges for
us to realize ground state cooling. On the other hand, since these conditions can also be difficult
to meet for other interesting systems like directly laser cooled molecules, the techniques we
use to overcome these challenges could be useful for a wider variety of systems.

As mentioned previously, in order to perform Raman sideband cooling efficiently and
minimize the heating during the optical pumping, we need to have a low initial temperature and
a small Lamb-Dicke parameter. However, since the Lamb-Dicke parameter $\eta$ is inversely
proportional to $\sqrt{m}$ (???) where $m$ is the mass of the atom, it is larger for sodium
given the same trap depth (???).
With $45mW$ of power at the focus of the trap, we measured a trapping frequency of
$\{\omega_1,\omega_2,\omega_3\}/2\pi = \{69, 430, 590\}\ \text{\text{kHz}}$
which corresponds to optical pumping Lamb-Dicke parameters of
$\{\eta^{OP}_1,\eta^{OP}_2,\eta^{OP}_3\} = \{0.60, 0.24, 0.21\}$.
Moreover, due to the small hyperfine structure in the $3^2P_{3/2}$ manifold,
the sub-Doppler cooling in sodium is also less efficient and we start the
Raman sideband cooling with a initial temperature of $40\mu K$. Combined with the high Lamb-Dicke
parameters, this gives us a initial effective optical pumping Lamb-Dicke parameters of
$\{\eta^{OP}_{1eff},\eta^{OP}_{2eff},\eta^{OP}_{3eff}\} = \{2.97, 0.52, 0.40\}$
As a result, there is a very high probability of heating during the optical pumping step
(figure \ref{f-ld}B) causing a $30\%$ average heating probability during optical pumping
in the weakly confined axial direction. Fortunately, the high Lamb-Dicke parameters also
provide us tools to overcome this issue. From our geometry, the Lamb-Dicke parameters for
Raman transitions are $\{\eta^R_{1},\eta^R_{2},\eta^R_{3}\} = \{0.40, 0.34, 0.29\}$. As shown in
figure \ref{f-ld}A, the high Raman Lamb-Dicke parameters causes a strong coupling to higher orders
of cooling sidebands, especially for high motional states.
This enables us to cool atoms in high motional states by driving on high order Raman sidebands,
removing more motional energy in a single cooling pulse and offsetting the effect of
strong heating. Since the heating probability is higher for high motional states,
cooling on these high order sidebands can greatly suppress the high heating during
optical pumping during the initial cooling. Since the coupling strength of different orders
do not reach their minimums at the same time, using multiple orders of motional sidebands
for cooling also avoids accumulation of population near the coupling minimum of a particular
order, which improves the overall efficiency of the cooling process.

@@

For the more tightly confined radial directions,
we starts the cooling with $\{\bar n_2, \bar n_3\}=2.2, 1.9$.
The radial sideband spectra of the initial distribution is shown in figure \ref{f-radial}A
where we can clearly see the first order heating, first order cooling and
second order cooling sidebands.
After applying about 1000 cooling pulses cooling in all three dimensions
starting with cooling on the radial second order,
the Raman spectrum with the same parameter is shown in figure \ref{f-radial}B,
where the first and second order cooling sidebands on both axis are suppressed.
Given the absence of the second order cooling sideband,
we can estimate the ground state probability in each direction based on the height of
the first order cooling and heating sideband to be $89.7(19)\%$ and $93.8(26)\%$.
Due to the complexity of the sideband structure,
we measured the Rabi flopping on the carrier and heating sideband before and after cooling
(\ref{f-radial}C and \ref{f-radial}D)
as a independent way to obtain the population in different motional states.
Fitting the Rabi flopping data give a ground state probability of $????$ and $????$,
showing good agreement between the two methods.

The axial direction is much less confined and therefore has a much higher initial effective
Lamb-Dicke parameter of $\eta^{OP}_{eff1}\approx3.3$.
The line in figure \ref{f-axial}A shows the Raman spectrum in the axial direction
before applying Raman sideband cooling where we can see clearly resolved Raman cooling sidebands
up to the eighth order suggesting that we have many motional states populated in this direction.
Therefore, we starts our cooling sequence by driving Raman transitions on the eighth order axial
cooling sideband. After applying the cooling sequence identical to the one we use to cool
the radial directions, the spectrum is shown in figure \ref{f-axial}B.
All of the high order ($\geqslant2$) axial cooling sidebands disappeared and the first order
cooling sideband is also strongly suppressed.
The ground state probability calculated from this spectrum is $????$.
We can also use Rabi flopping on the carrier and heating sideband to verify this result
similar to the radial direction. In this case, we see very good agreement on the carrier
(figure \ref{f-axial}B) but there is additional decoherence on the axial first order
heating sideband (figure \ref{f-axial}C).
We believe this decoherence is caused by technical noise in our setup which
produces the strongest effect on this transition due to its slow Rabi frequency.
The decoherence time scale agree with the magnetic field fluctuation we measured which produces
a Zeeman shift of around $5\text{kHz}$.

>>>>

\begin{figure*}
  \includegraphics[height=5.2cm]{imgs/Na_RSC_schematic.pdf}
  \includegraphics[height=4.5cm]{sequence.pdf}
  \caption{(A) Energy levels and schematics of Raman sideband cooling.
    The Raman transitions has a one photon detuning $\Delta=25GHz$ from the D2 line.
    We use D1 light with $\sigma^-$ polarization to repump atom out of $|F=2,m_F=-1\rangle$
    state to minimize heating on the atom in $|F=2,m_F=-2\rangle$ state.
    (B) Geometry and polarizations of the Raman and optical pumping beams relative to the
    optical tweezer and bias magnetic field.
    (C) Schematics of the cooling sequence. The tweezer is switching at $3\text{MHz}$ to
    reduce light shift during optical pumping. Each cooling cycle consists of $8$ pulses.
    The four axial pulses are addressing two neighboring cooling orders.
    The two pulses in each radial directions are either addressing two neighboring cooling orders
    or having different length on the first order when most of the population are below $n=3$
    towards then end of the cooling sequence.
    \label{f-setup}}
\end{figure*}
\begin{figure}[b]
  \includegraphics[width=8.5cm]{imgs/fig2_raman_op.pdf}
  \caption{Matrix elements and heating probabilities as a function of motional state.
    The range plotted covers $99\%$ of the initial thermal distribution.
    (A) Matrix elements for Raman transition in the axial direction showing deviation from
    $\sqrt{n}$ scaling and multiple minimums for different sideband orders.
    (B) Heating probability during the optical pumping step for all three axis.
    Due to the large Lamb-Dicke parameter,
    there is a high probability of heating especially in the axial direction.
    \label{f-ld}}
\end{figure}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_r.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_r3_p1.pdf}
  \caption{Radial Raman sideband spectrum of first order heating, first order cooling and
    second order cooling (A) before and (B) after Raman sideband cooling.
    Rabi flopping on axis 3 (C) carrier and (D) first order heating sideband
    after cooling.
    Solid lines in (C) and (D) are theoretical fit with a ground state probability of $93\%$.
    \label{f-radial}}
\end{figure*}
\begin{figure*}
  \includegraphics[height=4.2cm]{imgs/spectrum_a1.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_0.pdf}
  \includegraphics[height=4.2cm]{imgs/rabi_flop_a1_p1.pdf}
  \caption{Axial Raman sideband spectrum from first order heating to eighth order cooling
    (A) before and (B) after Raman sideband cooling.
    The data for the second and higher orders of cooling sidebands are taken with $150\mu s$
    pulse time and the rest are taken with $125\mu s$ pulse time.
    Rabi flopping on axial (C) carrier and (D) first order heating sideband
    after cooling.
    Solid lines in (C) and (D) are theoretical fit with a ground state probability of $92\%$.
    \label{f-axial}}
\end{figure*}

@@

The Raman sideband cooling we used to achieve the high ground state preparation fidelity
consists of multiple cycles of laser pulses to manipulate the internal and
motional states of the atoms.
Figure \ref{f-setup}A shows the schematic of the energy levels and the cooling sequence
in our setup.
Each cooling cycle starts with the sodium atom in the $|F=2, m_F=-2\rangle$
ground electronic state and a certain vibration state $n$.
In the first step, a Raman pulse drives a transition to the motional state $n-\Delta n$
to reduce the motional energy while also changing the internal state to $|F=1, m_F=-1\rangle$.
In the second step, which finishes the cooling cycle,
an optical pumping pulse bring the atom back to the $|F=2, m_F=-2\rangle$ state to take away
the entropy. The second step could also change the motional state of the atoms which
could cause heating. The possibility for this to happen for an atom in the motional level $n$
is approximately proportional to the effective Lamb-Dicke parameter
$\eta^{OP}_{eff}=\sqrt{2n+1}\eta^{OP}$ where $\eta^{OP}=k^{OP}\sqrt{\hbar/2m\omega}$ is the Lamb-Dicke parameter for
optical pumping.
The geometry of the relevant beams and their polarizations is showed in figure \ref{f-setup}B.
The optical tweezer has one weakly confined axial direction (axis $1$) and
two more strongly confined radial directions (axis $2$ and $3$).
Multiple pairs of beams are used to drive the Raman transition during cooling in order to
isolate and maximize the coupling to different trap axis.

Although Raman sideband cooling to motional ground state has already been successfully
implemented on neutral atoms in other experiments using heavier species like Rubidium
\cite{Thompson2013,Kaufman2012}
and Cesium [??], such experiments generally use good polarization gradient cooling
and a tight confinement to achieve a small Lamb-Dicke parameter and effective Lamb-Dicke parameter.
However, this regime is harder to achieve with sodium atoms, creating additional challenges for
us to realize ground state cooling. On the other hand, since these conditions can also be difficult
to meet for other interesting systems like directly laser cooled molecules, the techniques we
use to overcome these challenges could be useful for a wider variety of systems.

As mentioned previously, in order to perform Raman sideband cooling efficiently and
minimize the heating during the optical pumping, we need to have a low initial temperature and
a small Lamb-Dicke parameter. However, since the Lamb-Dicke parameter $\eta$ is inversely
proportional to $\sqrt[4]{m}/\lambda$ where $m$ is the mass of the atom and $\lambda$
is the wavelength of the atomic transition,
it is larger for sodium given the same trap depth due to its low mass and short D line wavelength.
With $45mW$ of power at the focus of the trap, we measured a trapping frequency of
$\{\omega_1,\omega_2,\omega_3\}/2\pi = \{69, 430, 590\}\ \text{\text{kHz}}$
which corresponds to optical pumping Lamb-Dicke parameters of
$\{\eta^{OP}_1,\eta^{OP}_2,\eta^{OP}_3\} = \{0.60, 0.24, 0.21\}$.
Moreover, due to the small hyperfine structure in the $3^2P_{3/2}$ manifold,
the sub-Doppler cooling in sodium is also less efficient and we start the
Raman sideband cooling with a initial temperature of $70\mu K$. Combined with the high Lamb-Dicke
parameters, this gives us a initial effective optical pumping Lamb-Dicke parameters of
$\{\eta^{OP}_{1eff},\eta^{OP}_{2eff},\eta^{OP}_{3eff}\} = \{3.97, 0.67, 0.50\}$
As a result, there is a very high probability of heating during the optical pumping step
(figure \ref{f-ld}B) causing a $30\%$ average heating probability during optical pumping
in the weakly confined axial direction. Fortunately, the high Lamb-Dicke parameters also
provide us tools to overcome this issue. From our geometry, the Lamb-Dicke parameters for
Raman transitions are $\{\eta^R_{1},\eta^R_{2},\eta^R_{3}\} = \{0.40, 0.34, 0.29\}$. As shown in
figure \ref{f-ld}A, the high Raman Lamb-Dicke parameters causes a strong coupling to higher orders
of cooling sidebands, especially for high motional states.
This enables us to cool atoms in high motional states by driving on high order Raman sidebands,
removing more motional energy in a single cooling pulse and offsetting the effect of
strong heating. Since the heating probability is higher for high motional states,
cooling on these high order sidebands can greatly suppress the high heating during
optical pumping during the initial cooling. Since the coupling strength of different orders
do not reach their minimums at the same time, using multiple orders of motional sidebands
for cooling also avoids accumulation of population near the coupling minimum of a particular
order, which improves the overall efficiency of the cooling process.

@@

For the more tightly confined radial directions,
we starts the cooling with $\{\bar n_2, \bar n_3\}=2.2, 1.9$.
The radial sideband spectra of the initial distribution is shown in figure \ref{f-radial}A
where we can clearly see the first order heating, first order cooling and
second order cooling sidebands.
After applying about 1000 cooling pulses cooling in all three dimensions
starting with cooling on the radial second order,
the Raman spectrum with the same parameter is shown in figure \ref{f-radial}B,
where the first and second order cooling sidebands on both axis are suppressed.
Given the absence of the second order cooling sideband,
we can estimate the ground state probability in each direction based on the height of
the first order cooling and heating sideband to be $89.7(19)\%$ and $93.8(26)\%$.
Due to the complexity of the sideband structure,
we measured the Rabi flopping on the carrier and heating sideband before and after cooling
(\ref{f-radial}C and \ref{f-radial}D)
as a independent way to obtain the population in different motional states.
Fitting the Rabi flopping data give a ground state probability of $90\%$ and $92\%$,
showing good agreement between the two methods.

The axial direction is much less confined and therefore has a much higher initial effective
Lamb-Dicke parameter of $\eta^{OP}_{eff1}\approx3.97$.
The line in figure \ref{f-axial}A shows the Raman spectrum in the axial direction
before applying Raman sideband cooling where we can see clearly resolved Raman cooling sidebands
up to the eighth order suggesting that we have many motional states populated in this direction.
Therefore, we starts our cooling sequence by driving Raman transitions on the eighth order axial
cooling sideband. After applying the cooling sequence identical to the one we use to cool
the radial directions, the spectrum is shown in figure \ref{f-axial}B.
All of the high order ($\geqslant2$) axial cooling sidebands disappeared and the first order
cooling sideband is also strongly suppressed.
The ground state probability calculated from this spectrum is $91.6(28)\%$.
We can also use Rabi flopping on the carrier and heating sideband to verify this result
similar to the radial direction. In this case, we see very good agreement on the carrier
(figure \ref{f-axial}B) but there is additional decoherence on the axial first order
heating sideband (figure \ref{f-axial}C).
We believe this decoherence is caused by technical noise in our setup which
produces the strongest effect on this transition due to its slow Rabi frequency.
The decoherence time scale agree with the magnetic field fluctuation we measured which produces
a Zeeman shift of around $5\text{kHz}$.
